A lottery model of density-dependent selection

### 2017-10-26 manuscript.tex
latest

MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at higher densities \citep{macarthur_1962,macarthur_1967}. While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
also initiated a sizeable literature on density-dependent selection. This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.

In MacArthur's view, selection at high densities is ``$K$-selection'', such that in ``crowded populations ... fitness is $K$''. MacArthur's argument is based on general equations for density-regulated population growth
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2),
\end{equation}
where, by definition, each

 $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.

%Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size. This led to the notion of ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equations \citep{gill_1974,case_1974,joshi_2001}.

Yet selection at high population density clearly involves more than carrying capacity.

and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].

>>>>

MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}. While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
also initiated a sizeable literature on density-dependent selection. This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.

Verbal interpretations of the $r$/$K$ scheme, and the logistic model, tend to obscure the generality of MacArthur's argument, and so it is worth briefly reviewing that argument here. Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
\end{equation}
where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$. By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not. This region locally corresponds to greater $n_1$ and/or $n_2$ than can be tolerated by type $1$, because $f_1$ and $f_2$ are strictly decreasing. In this sense, selection at high density favors the ability to keep growing at ever greater densities.

MacArthur then analyzed the possible outcomes of Eq.~\eqref{eq:macgeneral} in terms of "carrying capacities" (generally labeled ``$K$''), defined as the points where the nullclines intersect the axes (i.e. $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$). These intersection points neatly summarize the high density behavior of each type provided that the nullclines are approximately linear. He concluded that in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}. Note that only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types. For instance, if $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, then  $K_{12}=\alpha_{12}$ is just the standard Lotka-Volterra competition coefficient. This led to the distinction between selection for greater saturation density ($K$-selection in the strict sense) and selection for greater competitive ability (``$\alpha$-selection'') at high density \citep{gill_1974,case_1974,joshi_2001}.

Yet the $r$/$K$ scheme is strikingly incompatible with the  model of selection
